---
title: "311_calls"
format: html
editor: visual
---




```{r}
library(tidyverse)
library(janitor)
df <- read.csv("data/employeeearningscy18full.csv")
df<-clean_names(df)

```

You can add options to executable code like this

```{r}
#| echo: false
glimpse(df)
```

The `echo: false` option disables the printing of code (only output is displayed).

How many uniques? How many unique departments are in this payroll dataset?
```{r}
df |> count(department_name) |> 
  nrow()

```

What's the max per group? Who had the highest total earnings in the Boston Police Department and in the Parks Department?

```{r}
 df |> 
  filter(department_name %in% c("Boston Police Department", "Parks Department")) |> 
group_by(department_name) |> 
  slice_max(`total_earnings`, n = 1) |>
  select(name, department_name, `total_earnings`)
```

Out of all the categories of pay (REGULAR, RETRO, OTHER, OVERTIME, INJURED, DETAIL, etc), how much is the city paying in overtime?
```{r}
df |> 
  summarise(total_overtime = sum(o, vertime, na.rm = TRUE))


  
```


What's the average? Salary (REGULAR)?

```{r}
df |> 
  mutate(regular = as.numeric(regular)) |> 
  summarise(avg_regular_salary =
              mean(regular, na.rm = TRUE))

  
  
```


Which five departments have the most employees?
```{r}

df |> 
  count(department_name, sort = TRUE) |> 
  slice_head(n = 5)
```


What's the max total pay in the BPS Transportation Department?
```{r}
df |> 
  filter(department_name == "BPS Transportation") |> 
  slice_max(total_earnings,
            n = 1,
            with_ties = FALSE) |> 
  select(name, total_earnings)
```


What's the average pay for all Cab Monitors in the BPS Transportation Department?
```{r}
df |> 
  filter(department_name == "BPS Transportation") |> 
  filter(str_detect(str_to_lower(title), "cab monitor")) |> 
  summarise(
    avg_total_pay =
      mean(total_earnings, na.rm = TRUE),
    employees = n()
  )
```


What's the difference in between average overtime by department? Between the fire department and boston police?
```{r}
df |> 
  mutate(overtime = as.numeric(gsub(",", "", overtime))) |> 
  filter(department_name %in% c("Boston Fire Department", "Boston Police Department")) |> 
  group_by(department_name) |> 
  summarise(avg_overtime = mean(overtime, na.rm = TRUE), .groups = "drop") |> 
  pivot_wider(names_from = department_name, values_from = avg_overtime) |> 
  mutate(difference = `Boston Fire Department` - `Boston Police Department`)
```

How much do employees with the last name "O'Malley" make in Boston together?
```{r}
df |> 
  filter(str_detect(name, "O'Malley")) |> 
  summarise(
    total_omalley_earnings = sum(total_earnings, na.rm = TRUE),
    employees = n()
  )
```


The example data set is for Boston payroll in 2018. If in 2020, Boston Police Department's total earnings was $512,399,200, what is the percent change from their total earnings in 2018?

```{r}
bpd_2018 <- df |> 
  filter(department_name ==
           "Boston Police Department") |> 
  summarise(total =
              sum(total_earnings,
                  na.rm = TRUE)) |> 
  pull(total)

bpd_2020 <- 512399200

percent_change <-
  ((bpd_2020 - bpd_2018) / bpd_2018) * 100

percent_change
```

