---
title: " Traffic enfoecement camera in Washington, D.C. distribution and revenue"
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
    code-summary: "Show the code"
    code-tools: true
execute:
  warning: false
  message: false
editor: visual
---

# Memo brief

this memo explore two main ideas about the automated camera enforcement in Washington, D.C.

1- Revenue v. safety: how much money the city actually billed in traffic enforcement and are they prioritizing safety or revenue.

2- Racial distribution: are wards with majority black population more likely to experience more automated enforcement.

## Data sources:

-   MPD arrest data for 2024

-   DC automated traffic enforcement locations

-   DC automated camera violations by month (2024)

-   ACS 5-year ward demographics

-   2024 moving violations

```{r}
library(tidyverse)
library(janitor)
library(lubridate)
library(sf)
library(ggrepel)
arrests <- read_csv("data/2024_dc_arrests.csv") |> clean_names()
demographics <- read_csv("data/ACS_5-Year_Demographic_Characteristics_DC_Ward.csv") |> clean_names()
social <- read_csv("data/ACS_5-Year_Social_Characteristics_DC_Ward.csv") |> clean_names()
camera_locs <- read_csv("data/Automated_Traffic_Enforcement.csv") |> clean_names()
camera_violations <- read_csv("data/Automated_Safety_Cameras_Violation_Count_By_Month.csv") |> clean_names()
# racial distribution
ward_pop <- demographics |> 
  mutate(ward_num = str_extract(namelsad, "\\d")) |> 
  select(ward_num, 
         total_pop = dp05_0001e, 
         black_pop = dp05_0038e,
         white_pop = dp05_0037e)
#education
ward_social <- social |> 
  mutate(ward_num = str_extract(namelsad, "\\d")) |> 
  select(ward_num, 
         total_households = dp02_0001e,
         pop_25_plus = dp02_0067e, 
         bachelors_plus = dp02_0068e)

# creating a table for each ward with total arrests and numbers for each racein this one I have to map areests psa to wards so it can match each ward in the district
ward_arrest_stats <- arrests |> 
  mutate(psa = as.character(offense_location_psa)) |> 
  mutate(ward_num = case_when(
    psa %in% c("301", "302", "303", "304", "305", "408") ~ "1",
    psa %in% c("101", "102", "207", "208", "209", "306", "307") ~ "2",
    psa %in% c("201", "202", "203", "204", "205", "206") ~ "3",
    psa %in% c("401", "402", "403", "404", "405", "406", "407", "409") ~ "4",
    psa %in% c("501", "502", "503", "504", "505", "506", "507") ~ "5",
    psa %in% c("103", "104", "105", "106", "107", "108") ~ "6",
    psa %in% c("601", "602", "603", "604", "605", "606", "607", "608") ~ "7",
    psa %in% c("701", "702", "703", "704", "705", "706", "707", "708") ~ "8",
    TRUE ~ NA_character_
  )) |> 
  group_by(ward_num) |> 
summarise(
    total_arrests = n(),
    black_arrests = sum(defendant_race == "BLACK", na.rm = TRUE),
    white_arrests = sum(defendant_race == "WHITE", na.rm = TRUE),
    total_traffic_arrests = sum(arrest_category %in% c("Traffic Violations", "Driving/Boating While Intoxicated"), na.rm = TRUE), 
    black_traffic_arrests = sum(arrest_category %in% c("Traffic Violations", "Driving/Boating While Intoxicated") & defendant_race == "BLACK", na.rm = TRUE), 
    white_traffic_arrests = sum(arrest_category %in% c("Traffic Violations", "Driving/Boating While Intoxicated") & defendant_race == "WHITE", na.rm = TRUE)
  )
#Traffic cameras violations
total_violations <- camera_violations |> 
  group_by(enforcement_space_code) |> 
  summarise(total_violations = sum(num_violations, na.rm = TRUE))
#Traffic cameras by ward 
ward_camera_stats <- camera_locs |> 
  mutate(ward_num = as.character(as.integer(ward))) |> 
  left_join(total_violations, by = "enforcement_space_code") |> 
  group_by(ward_num) |> 
  summarise(
    camera_count = n(),
    speed_camera_count = sum(enforcement_type == "Speed", na.rm = TRUE),
    total_camera_violations = sum(total_violations, na.rm = TRUE)
  ) |> 
  filter(ward_num != "0")
wards <- st_read("data/Wards_dc.geojson", quiet = TRUE) |>
  mutate(WARD = as.numeric(WARD)) |>
  st_transform(4326)
#now I will join the data and calculate the per capita rates
joined_df <- ward_arrest_stats |> 
  drop_na(ward_num) |> 
  inner_join(ward_pop, by = "ward_num") |> 
  left_join(ward_camera_stats, by = "ward_num") |> 
  mutate(
    # Arrests per 1,000 residents of each race (per capita)
    black_per_capita = (black_arrests / black_pop) * 1000,
    white_per_capita = (white_arrests / white_pop) * 1000,
    #Cameras per 10,000 residents
    cameras_per_10k = (camera_count / total_pop) * 10000,
    # Percentages for dataviz
  pct_black_pop = (black_pop / total_pop) * 100,
    pct_white_pop = (white_pop / total_pop) * 100,
    pct_black_arrests = (black_arrests / total_arrests) * 100,
    pct_white_arrests = (white_arrests / total_arrests) * 100,
    pct_black_traffic_arrests = (black_traffic_arrests / total_traffic_arrests) * 100,
    pct_white_traffic_arrests = (white_traffic_arrests / total_traffic_arrests) * 100

  )
# Loading 2024 monthly revenue
jan <- read_csv("data/Moving_Violations_Issued_in_January_2024.csv") |> clean_names()
feb <- read_csv("data/Moving_Violations_Issued_in_February_2024.csv") |> clean_names()
mar <- read_csv("data/Moving_Violations_Issued_in_March_2024.csv") |> clean_names()
apr <- read_csv("data/Moving_Violations_Issued_in_April_2024.csv") |> clean_names()
may <- read_csv("data/Moving_Violations_Issued_in_May_2024.csv") |> clean_names()
jun <- read_csv("data/Moving_Violations_Issued_in_June_2024.csv") |> clean_names()
jul <- read_csv("data/Moving_Violations_Issued_in_July_2024.csv") |> clean_names()
aug <- read_csv("data/Moving_Violations_Issued_in_August_2024.csv") |> clean_names()
sep <- read_csv("data/Moving_Violations_Issued_in_September_2024.csv") |> clean_names()
oct <- read_csv("data/Moving_Violations_Issued_in_October_2024.csv") |> clean_names()
nov <- read_csv("data/Moving_Violations_Issued_in_November_2024.csv") |> clean_names()
dec <- read_csv("data/Moving_Violations_Issued_in_December_2024.csv") |> clean_names()
#combine all 12 in 1 table

tickets_revenue_2024 <- bind_rows(
  jan, feb, mar, apr, may, jun,
  jul, aug, sep, oct, nov, dec
)

```

# Background

## Physical traffic violation arrests vs. automated enforcement arrests

this section compares the physical arrests vs. automated enforcement pattern for each ward

```{r}
#| echo: false
#| warning: false
joined_df |> 
  select(ward_num, pct_black_pop, pct_black_arrests, pct_white_pop, pct_white_arrests) |> 
  pivot_longer(cols = starts_with("pct"), names_to = "metric", values_to = "value") |> 
  mutate(ward_num = factor(ward_num, levels = as.character(1:8))) |> 
  ggplot(aes(x = ward_num, y = value, fill = metric)) +
  geom_col(position = "dodge") +
  # data labels
  geom_text(aes(label = paste0(round(value, 0), "%")), 
            position = position_dodge(width = 0.9), 
            vjust = -0.5, size = 2) +
  # assigning colors (I thought these colors will be eaasy to distonguish but let me know if they are too much)
  scale_fill_manual(values = c(
    "pct_black_pop" = "green",      
    "pct_black_arrests" = "red",    
    "pct_white_pop" = "grey",       
    "pct_white_arrests" = "black"), 
    labels = c("Black Residents", "Black Arrests", "White Residents", "White Arrests")
  ) +
  theme_minimal() +
  # editing the graph by removing background abd y axis
  theme(
    panel.grid = element_blank(),
    axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    legend.position = "top"
  ) +
  labs(title = "Residents vs. Arrests: Racial Representation by Ward",
       x = "Ward", fill = "",
       caption = "Source: Metropolitan Police Department (2024 Arrest Data) & U.S. Census Bureau (ACS 5-Year)")
```

For more background context the table below shows how many time more likely a black resident is arrested then a white resident in each ward.

```{r}
joined_df |> 
  mutate(disparity_ratio = black_per_capita / white_per_capita) |> 
  select(ward_num, black_per_capita, white_per_capita, disparity_ratio) |> 
  arrange(desc(disparity_ratio))
```

# Camera distribution across wards

After exploring the physical traffic arrests, the next step is to explore where the automated enforcement cameras are distributed around the district.

By counting the number of cameras in each ward as well as the volume and the revenue of the traffic enforcement we can get a clear understanding whether certain neighborhoods are experiencing more traffic enforcement.

## Traffic cameras by ward

```{r}
ward_camera_stats |> 
  select(ward_num, camera_count, total_camera_violations) |> 
  arrange(desc(camera_count))

```

The Table above shows that ward 5, 7 and 8 have the highest number of traffic camera enforcement as well as the highest number of violations.

## Enforcement camera density in each ward

this chart compares the camera density in each ward using camera per 10,000 resident.

```{r}
#| echo: false
#| warning: false
ggplot(joined_df, aes(x = reorder(paste("Ward", ward_num), cameras_per_10k), y = cameras_per_10k, fill = pct_black_pop)) +
  geom_col() +
  coord_flip() +
  geom_text(aes(label = round(cameras_per_10k, )), hjust = -0.2, size = 2.3) +
  scale_fill_gradient(low = "lightgrey", high = "darkblue") +
  theme_classic() +
  theme(
    axis.line = element_blank(),
    axis.ticks = element_blank(),
    axis.text.x = element_blank(),
    axis.title.x = element_blank(),
    axis.line.y = element_blank()
  ) +
  labs(title = "Traffic Camera Density by Ward",
       subtitle = "Number of cameras per 10,000 residents.",
       x = "", y = "Cameras per 10k Residents", fill = "% Black Population",
       caption = "Source: Open Data DC (Automated Traffic Enforcement)")
```

## Traffic violation volume by neighborhood

```{r}
#| message: false
#| warning: false

library(sf)
library(tidyverse)
library(lubridate)

# Violation data for 2024
top_locations <- camera_violations |> 
  mutate(date = ymd_hms(year_month)) |> 
  filter(year(date) %in% c(2024)) |> 
  group_by(enforcement_space_code) |> 
  summarise(total_violations = sum(num_violations, na.rm = TRUE))

# perparing camera coordinates to add them to the map
camera_locs_clean <- camera_locs |>
  rename_with(tolower) |>
  rename(
    longitude = any_of(c("longitude","long","x","lon")),
    latitude  = any_of(c("latitude","lat","y"))
  )

# converting camera location to special points
camera_dots_sf <- camera_locs_clean |>
  drop_na(longitude, latitude) |> 
  st_as_sf(coords = c("longitude", "latitude"), crs = 3857) |>
  st_transform(4326)

# Loading DC neighborhood map
neighborhood_sf <- st_read("data/Neighborhood_dc.geojson", quiet = TRUE) |> 
  st_transform(4326)
neighborhood_sf <- neighborhood_sf |> 
  rename_with(tolower) |> 
  rename(neigh_name = any_of(c("name")))

# linkimg violation total and cam locations
camera_points_with_violations <- camera_dots_sf |> 
  left_join(top_locations, by = "enforcement_space_code")
camera_neighborhood <- st_join(camera_points_with_violations, neighborhood_sf)

# calculating total violation in each neighborhood
neighborhood_revenue <- camera_neighborhood |> 
  filter(!is.na(neigh_name)) |> 
  st_drop_geometry() |> 
  group_by(neigh_name) |> 
  summarise(
    total_revenue_vol = sum(total_violations, na.rm = TRUE)
  )
map_data_neighborhood <- neighborhood_sf |> 
  left_join(neighborhood_revenue, by = "neigh_name")

# drawing the map
ggplot() +
  geom_sf(data = map_data_neighborhood,
          aes(fill = total_revenue_vol),
          color = "white",
          linewidth = 0.2) +
  geom_sf(data = camera_dots_sf,
          color = "orange",
          size = 0.4,
          alpha = 0.4) +
  
  # shading
  scale_fill_viridis_c(
    option = "mako",
    direction = -1,
    name = "Ticket Volume",
    labels = scales::comma,
    na.value = "grey90"
  ) +
  
  theme_void() +
    labs(
    title = "Ticket volume by neighberhood",
    caption = "Source: Open Data DC"
  )
```

## Traffic enforcement violation and Black population

This scatter plot compares the number of tautomated traffic enforcement violations with the percentage of black population in each ward using the number of cameras per 10,000 resident.

```{r}
#| message: false
#| warning: false

ggplot(joined_df, aes(x = pct_black_pop, y = cameras_per_10k)) +
  
  # formating the point size
  geom_point(aes(size = total_camera_violations),
             color = "darkblue",
             alpha = 0.8) +
  
  # labeling each ward
  geom_text(aes(label = paste("Ward", ward_num)),
            nudge_y = 0.1,
            size = 3) +
  
  # adding the line to emphasize relation ship 
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  
  theme_minimal() +
  
  scale_x_continuous(labels = scales::percent_format(scale = 1)) +
  
  labs(
    title = "Traffic CamerasViolations vs Black Population by Ward",
    subtitle = "Point size represents total automated traffic violations",
    x = "% Black Residents",
    y = "Cameras per 10,000 residents",
    size = "Total Camera Violations",
    caption = "Source: Open Data DC & U.S. Census Bureau"
  )
```

# Revenue generated by traffic enforcement

This section explores the amount of automated enforcement fines in the district and how much money was actually paid

## Total Revenue

The total revenue that the district should get from traffic enforcement is \$341,264,692

```{r}
tickets_revenue_2024 |> 
  summarise(
    total_fines = sum(fine_amount, na.rm = TRUE)
  )
```

## Collected revenue

Less then the third of the total revenue was paid to the District.

```{r}

tickets_revenue_2024 |> 
  summarise(
    fines = sum(fine_amount, na.rm = TRUE),
    paid = sum(total_paid, na.rm = TRUE),
    payment_rate = (paid / fines) * 100
  )
```

## Revenue trend in 2024

## Top 5 camera generating revenue

The analysis shows that the top cameras generating revenues on main avenues and freeways.

```{r}

#| message: false
#| warning: false

library(sf)

ticket_points <- tickets_revenue_2024 |> 
  filter(!is.na(longitude) & !is.na(latitude)) |> 
  mutate(
    longitude = as.numeric(longitude),
    latitude = as.numeric(latitude)
  ) |> 
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326, remove = FALSE)

ticket_points_geo <- ticket_points |> 
  st_join(
    wards |> 
      mutate(ward_num = as.character(WARD)) |> 
      select(ward_num)
  )

top5_cameras_revenue_2024 <- ticket_points_geo |> 
  st_drop_geometry() |> 
  group_by(location, ward_num) |> 
  summarise(
    tickets_issued = n(),
    total_fines = sum(fine_amount, na.rm = TRUE),
    total_paid = sum(total_paid, na.rm = TRUE),
    payment_rate = (total_paid / total_fines) * 100,
    .groups = "drop"
  ) |> 
  arrange(desc(total_fines)) |> 
  slice_head(n = 5) |> 
  select(location, ward_num, tickets_issued, total_fines, total_paid, payment_rate)

top5_cameras_revenue_2024
```

## Monthly revenue trend in 2024

this line chart shows that the most months generating revenue are between April and July and go down in august. Although the trend go back up again in September but it doesn't reach the same levels as April- July.

```{r}
#| message: false
#| warning: false

revenue_trend <- tickets_revenue_2024 |> 
  mutate(issue_date = as.Date(issue_date),
         month = floor_date(issue_date, "month")) |> 
  group_by(month) |> 
  summarise(total_revenue = sum(total_paid, na.rm = TRUE))

ggplot(revenue_trend, aes(x = month, y = total_revenue)) +
  geom_line(color = "darkred", linewidth = 1) +
  theme_minimal() +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Traffic Ticket Revenue Trend 2024",
    subtitle = "Monthly revenue collected from moving violations",
    x = "Month",
    y = "Revenue Collected in $",
    caption = "Source: Open Data DC"
  )
```
